A heterogeneous landscape of selection and epistasis in protein-coding genes revealed by two-locus statistics

### 2022-03-31 response_to_reviewers/response.tex

Your manuscript has been evaluated by two reviewers.  I agree with the
assessment of the reviewers of that this has the potential to be an important
contribution but that it is not there yet.  Your manuscript would need to be
revised substantially, including conducting additional analysis, to be
acceptable for publication in GENETICS.  Please understand that incremental
changes will not be sufficient.  If you intend to resubmit, please let the
editorial office know approximately how long you expect to need for revisions.
Upon resubmission, please include a detailed response to the reviewers'
comments and to the concerns listed above; it will likely be sent out for
review.  I am hopeful you will be willing to do these revisions as I am excited
about the potential of your manuscript.

\begin{response}
    Dear Dr. Agrawal,

    Thank you for the opportunity to revise this manuscript for consideration
    in Genetics and your patience in completing the revisions. I appreciate
    the careful reading and detailed comments that you and each reviewer have
    made, and I believe that these suggestions have substantially improved the
    paper. Below, my responses to each comment are indented and in blue.

    I'd first like to highlight a few of the more substantial changes:
    \begin{itemize}
        \item The addition of a large suite of numerical validations against
            individual-based simulations, as well as a handful of forward
            simulations to demonstrate the effects of many linked loci on diversity
            statistics
        \item A new Figure 1 that provides some intuition for the two-locus
            sampling distribution and its summaries (contrasted with the SFS
            and its summaries)
        \item Extended numerical analyses, in particular for patterns of LD under
            dominance models
        \item Partitioning LD by allele-frequency binning -- in particular, the
            result of increased LD among missense mutations within conserved
            domains is driven by common variants above $\approx 10\%$
            (I expand on this below)
    \end{itemize}

@@

The manuscript has a theoretical and empirical component and each needs
revision.  The reviewers have each made a number of comments that should be
addressed and I will summarize the main ones from my perspective.

Theoretical Section: I agree with the reviewers that there is value in your
numerical approach to two-locus sampling distribution. This comment from
Reviewer 2 captures my sentiment: ``the first part of the manuscript does not
seem to fully capitalize on the potential that these new numerical approaches
offer to build quantitative in the extremely large space of models that is
explored.'' Here are what I view as the most important ways the theory section
of the ms should be developed:

1) The numerical approach is based on some approximations but there is no
testing of how well these approximations hold. I agree with both reviewers that
there should be some level of testing (via comparison to forward simulations)
to show how well these approximations work and, when they fail, what kind of
biases they cause.

\begin{response}
    The Supporting Material now includes validation using discrete
    forward simulation. The forward simulations are individual based
    Wright-Fisher simulations, with implementation details in the Supporting
    Material. These comparisons show that the numerical approach introduced in
    this paper is accurate for the weak to moderate recombination and selection
    strengths studied in this paper ($\rho, \gamma \lesssim 30$).
    Most simulations only tracked the allele frequencies at the two loci,
    neglecting the background selection-like effects,
    although I also ran a small number of simulations of larger genomic
    regions that allowed many mutations to segregate at once. These are included
    in the Supporting Information.
\end{response}

@@

3) Both reviewers pointed out that selection occurs at more than two sites and
that pairwise LD could be affected by selection at other loci. That seems an
important caveat that should be stated and ideally tested.

\begin{response}
    Thank you for the suggestion. I now include this caveat at the beginning
    of the results. I also include a some simulations of large regions with
    many selected loci to give a sense of the bias caused by selection at other
    loci than the focal pair.
\end{response}

4) I did not look at the scripts you made available but I would encourage you
to make them as user-friendly as possible. This ms has the potential to make a
bigger impact if it is easy for empirical (not just theoretical!) population
geneticists to use your scripts.

\begin{response}
    I am happy to make all scripts public, and I have tried to keep
    an organized an documented/commented repository of all scripts used in the
    analysis here (\url{https://www.github.com/apragsdale/two_locus_selection.git}).
    Additionally, the numerical methods are implemented, documented, and tested
    within the \emph{moments} software (which includes single-locus SFS and
    multi-population LD methods, documentation at \url{https://moments.readthedocs.io/}),
    which I and other developers actively maintain.
\end{response}

@@

    In interpreting this result, this suggests to me that there is something
    more interesting going on than the explanation that common missense
    mutations are simply effectively neutral. Assuming synonymous mutations are
    effectively neutral (this is of course not always the case, although common
    synonymous mutations would be neutral by the same argument, and we don't
    see a difference in Tajima's D for syn variation within vs outside of domains),
    we still observe increased LD of missense mutations over that of synonymous
    mutations. If missense mutations at these common frequencies were entirely
    neutral, we'd expect their levels of LD
    to be roughly equal to those of synonymous mutations.

    Background selection also appears to have minimal effects on signed LD between
    neutral mutations (Supp Figures 7--10), so I lean toward the explanation that
    some form of direct selection/interactions on those missense mutations (or at
    least a subset of them) is driving this signal. More careful analysis is needed
    to say anything definitively, which I have left for future work.
\end{response}

6) Distances. The average pairwise distance will be lower for mutational
classes that are more abundant. Presumably LOFs are much less abundant than
synonymous or missense mutations so I suspect that the average pairwise
distance between LOFs is larger than for other two types (which probably differ
from one another). Though this won't affect the sign of LD it is obviously
expected to affect the magnitude.

\begin{response}
    In the previous version of the manuscript, I had considered average LD for
    pairs of mutations outside of domains weighted to match the distances between
    mutations observed inside domains. I repeated that analysis here and placed
    more emphasis on comparing pairs at matching distances (see Supp Fig 26 for a
    comparison of density of pairs in differnece classes at given distances).
    For the LOFs, that is a good point, although how few LOF pairs exist within
    the same gene makes further subsetting of LOFs difficult.
    I now include this caveat in the Discussion.
\end{response}

@@

\begin{response}
    My apologies for potentially misrepresenting the results from those simulations
    in Sandler et al -- that certainly wasn't my intention. I have removed these
    lines in this version of the manuscript, though I do think the observation is
    important that positive LD between synonymous mutations only extends over short
    distances (a few hundred base pairs). I have edited the Discussion about this point.
\end{response}

12) Line 350-360. The point about multinucleotide mutations (MNMs) is
important. I would have liked to see more detail in the legend of Figure S8 of
what was actually done to produce this figure (i.e., enough of a methods
description that one could reproduce this). While the model fits reasonably
well, it implicates what seems like a very proportion of multinucleotide
mutation events. Is there any evidence to support this, perhaps from Icelandic
trio dataset? I am not suggesting you need to do a formal analysis (though it
would be great if you did) but you should comment on whether there is any
evidence that the estimated fraction of MNMs is realistic. (A personal note: I
think you are right to point out that MNMs make the ``null'' expectation to be
positive LD because MNMs are a biological reality. I take the same attitude to
population structure / gene flow. The question then becomes about the relative
importance of the two as the source of positive LD for neutral variants. I do
not claim to know with certainty the answer to that question but I am highly
skeptical that MNMs are sufficient to explain long range positive LD. To be
clear, I am not requiring you to convince me otherwise, but I will be impressed
if you can!)

\begin{response}
    I have extended the methods description for this analysis in the Supplement, as
    well as refined the inference to directy estimate the proportion of mutations
    that result in MNM events instead of simply fitting the probability that an
    \emph{observed} pair of mutations at a given distance were caused by a MNM event.

    In the previous version of the manuscript, I pointed out (though I think not too
    clearly) that the proportion of pairs of mutations observed at a given distance
    that arose through a MNM event is different than the proportion of new mutation events
    that are single- vs multi-nucleotide mutations. The probability that a new
    mutation results in a MNM can be very small (much less than 1\%), but the proportion
    of mutation pairs we observe that were initially MNM events can be quite high
    (e.g., 5--10\%). I expand on this point in the Supplement.

@@

    I also now include a few forward-in-time individual-based simulations
    of large regions with high rates of selected mutations, to get an idea of the
    magnitude of distortion that can occur due to background selection, associative
    overdominance, and linked purifying selection. These are included in the Supplement.
\end{response}

2) It would be good if some of the predictions for the behavior of LD
statistics in complex models could be verified through forward in time
simulations (using SLiM or something similar). This isn't necessary for all
models (because that is the point of the numerical methods to avoid lots of
simulations), but I think it would add rigor to this work to show that at least
for some models, the predictions match up to the simulations.

\begin{response}
    I have now included a simulation of a region allowing many simultaneous
    deleterious mutations (with only Hill-Robertson interference, that is,
    no epistasis), for additive ($h=0.5$) and recessive ($h=0, 0.1$) mutations.
    The goals of these larger simulations were to explore the effects of multiple
    linked selected loci.

    In addition to these simulation, I ran a large suite of discrete simulations
    tracking haplotype frequencies for pairs of loci under range of epistatic
    and dominance interaction scenarios, comparing the resulting two-locus
    sampling distribution. These are shown in the Supplement and online
    documentation of moments and this manuscript, and show good agreement between
    simulations (in the absense of BGS) and the numerical approach introduced here.
\end{response}

@@

3) Analysis of empirical data (Fig 4 and Fig 5): Missense and variants within
protein domains seem to be in more positive LD than those not within protein
domains or synonymous SNPs. This is an interesting and potentially important
result. It does not appear that the different types of variants were matched
for allele frequency. Missense and synonymous SNPs have very different allele
frequency distributions, with the former having more rare variants. It is
well-known that allele frequencies can influence LD statistics (see for
example, vanLiere \& Rosenberg (2008 TPB); Hedrick (1987 Genetics); Eberle et
al. (2006 PLoS Genetics)). Further, the fact that Table S4 shows that Tajima's
D differs among these different types of variants further supports the notion
that the allele frequencies inside protein domains differ from those outside.
Thus, I'm concerned that the differences in LD patterns can be simply driven by
differences in allele frequency, rather than differences in LD driven by
epistasis or compensatory mutations. Matching on allele frequency when
computing the LD statistics would alleviate this problem.

\begin{response}
    Thank you for these suggestions. While I had matched pairs of mutations based
    on distances, I had not matched by allele frequences. As mentioned above,
    I now condition on both alleles being rare, uncommon, or common and compare
    LD across these categories.
    From this, the opposing signals of increased/decreased
    LD between missense mutations within/outside of domains is driven by
    mutations with frequencies greater than $0.1$. The agreement between
    high-frequency-conditioned and unconditioned $\sigma_d^1$ is unsurprising,
    since $\sigma_d^1$ is dominated by mutations at common frequencies (while
    $r$ and $r^2$ are more sensitive to rare variants).
    I now include a much larger set of conditioned LD comparisons (in the
    Supplement, Tables S6--S11 and Figures S20-S24).
\end{response}

@@

\begin{response}
    Thank you for the suggestions. I have added references in the first paragraph
    of the Introduction as suggested.
\end{response}

3) The terms ``synergistic'' and ``antagonistic'' epistasis are used extensively
throughout. It would be good to provide a more clear definition of these terms
early on in the manuscript and how they are used here for describing
deleterious mutations.

\begin{response}
    Thank you for the suggestion. In the first paragraph of the Empirical observations
    section, where the term ``synergistic'' epistasis is first used, I attempted
    to make the definition a bit more clear. I do the same for both synergistic and
    antagonistic epistasis when introducing the epistasis models in the results.
\end{response}

4) Caption of Figure 1: I believe that the signs of the synergistic and
antagonistic epistasis are reversed. Given the definition in Methods, shouldn't
epsilon be $>0$ for synergistic epistasis as more deleterious mutations have
lower fitness than predicted under multiplicative models? Then, epsilon would
be $<0$ for antagonistic epistasis.

\begin{response}
    Thank you for catching this error, and I'm sorry if it caused any confusion.
    Caption fixed.
\end{response}

@@

5) Figures 1-2: I think it would be good to highlight that these plots show the
expected values of these LD statistics for the given models.

\begin{response}
    I have updated the figure captions to clarify that
    expectations of LD that are being plotted.
\end{response}

@@

7) Line 146: Cite some references if saying that the effect of epistasis on LD
is widely studied.

\begin{response}
    With the rearrangement of the text, this line has been removed.
\end{response}

8) Supplementary Figure 1: I think it would be instructive to move this to the
main text to show a more intuitive depiction of the 2-locus sampling
distribution.

\begin{response}
    A new Figure 1 in this version hopefully provides a more intuitive depiction
    of the two-locus sampling distribution and its summaries (with direct comparison
    to the SFS and some of its summaries). Thank for you this suggestion, and I hope
    the new figure is helpful in introducing these statistics.
\end{response}

@@

9) Lines 285-289: These lines seem at odds with what Garcia \& Lohmueller
reported. They found that interference interactions are likely common among
nonsynonymous mutations and that nonsynonymous variants have more LD than
synonymous ones.

\begin{response}
    Thanks for pointing this out, and this is now clarified in the Discussion.
\end{response}

@@

Major comments:

1) As far as I can tell, the calculations in the SI are correct under the
approximations made, but there is very little discussion of the importance of
the approximations made and the limitations these impose for interpreting data.
Specifically, the condition that $r$ and $s$ are order $1/N_e$ and that $n \ll
N_e$ implies that $ns \ll 1$. This means that this approach is unable to
accurately capture the statistics of mutations segregating at frequencies of
order $1/Ns$ or larger. I find this both broadly and specifically concerning.
Specifically, Good (2020) bioRxiv [cited within the discussion] has
demonstrated that it is precisely these mutations that show an excess of LD
that resembles the effects of antagonistic epistases, even in purely additive
models, which is not seen here. This concern is also directly related to the
finding that antagonistic epistasis may be responsible for unusually high LD
within genes -- it is unclear to me whether negative selection alone may be
responsible for these patterns.

\begin{response}
    I appreciate the concern about accuracy of the numerical methods, and I
    have added extensive validations against simulations for the cases
    of additivity, synergistic epistasis, antagonistic epistasis, and recessive
    mutations. For the parameter values considered here
    ($\rho, |\gamma| \approx 0 - 30$), we find very good agreement between simulation
    and numerical solutions. The condition that $r$ and $s$ are of order $1/N_e$
    is standard in the diffusion approximation, allowing us to neglect terms
    of order $s^2$, $rs$, $s/N_e$, etc. In practice, diffusion methods still
    provide reasonably accurate results for larger selection and recombination
    values, and only break down for very strong
    selection beyond the parameter ranges I considered here
    (see, e.g., Krukov and Gravel, bioRxiv 2021)

@@

    I don't exactly follow the conclusion, that ``This means that this approach
    is unable to accurately capture the statistics of mutations segregating at
    frequencies of order $1/Ns$ or larger.'' The condition that $ns\ll 1$ is only
    violated for extremely strong selection (say, $|s| \gtrsim 0.01$, as sample sizes
    that I consider here are not too large). But that strength of selection is beyond
    the results presented in this paper.

    I have expanded on the parameter ranges that this method is applicable for
    and point to simulation results showing accuracy over this range of parameters.
\end{response}

@@

2) More broadly, without more detailed exploration of the parameter space of
the models considered and without more work to interpret quanitatively at which
frequencies different behaviors will occur as a function of the parameters, it
is difficult to interpret the generalizability of the results quoted here. Note
that this is especially a concern when the results are used to shape
qualitative insights about the patterns seen in genomes. It would be important
to describe the theoretical reasoning behind the choice of specific values of
parameters the simulations, and why the author believes they tile the space of
important behaviors of a two locus model.

\begin{response}
    I agree that it is difficult to make more definitive claims about the
    causes of variation in patterns of LD seen in the data without a more
    detailed exploration of the relevant parameter space. Ideally, this would
    come in the form of likelihood-based inference of the joint effects of 
    selection and interactions, as well as validation against models of BGS.
    I comment on this in the Discussion but leave it to future work.

@@

    I have expanded the theoretical/numerical results to consider additional
    parameters and the dependence on frequency conditioning, though this still
    falls short of a complete exploration of parameter space. I focused primarily
    on selection strengths in the range $\gamma\approx -1$ to $-10$, as these
    selection strengths show the largest effects of interference without 
    epistasis, including for dominance. The very large space of possible selection
    models makes it difficult to fully tile this space. Nonetheless, I
    believe that the results presented in the main text and the expanded results
    in the Supplement are useful for building intuition about the behavior or
    two-locus statistics under different interaction models.
    I have also edited the Discussion to include these caveats and
    point to future work in these directions.
\end{response}

3) On a related note, given the restriction that $n \ll N$, many of these
results should be directly comparable to the recent analytical results from
Good (2020) bioRxiv on expected patterns of LD in a two-locus model. I think
that using the rationale developed by that and similar studies would really
make it possible to capitalize on the ability to rapidly calculate LD in a
broader range of contexts.

\begin{response}
    This is a good suggestion, I spent quite a bit of time trying to implement
    the analytical results from Good (2020). I was unable to make much
    progress in such a comparison. For the most part, I was able to replicate
    the results in Figures 3 and 4 in that paper (e.g., see below, though there
    are noticeable differences between the written formulae and the plots in
    Figure 3 in Good).
    But to compare the analytical results in Good (2020) directly
    to the numerical approximations
    here, we need to make use of the finite sample estimators (Appendix G in Good).
    Unfortunately, I was unable to obtain sensical results implementing the
    formulae in Appendix G, and there may be one (or more) typos in Appendix G
    that make a direct comparison impossible without rederiving those results.

@@

2) Figure 1: I suggest considering using different colors for panels C-F
compared to A-B to highlight that a different parameter is being varied in
these panels.

\begin{response}
    I have done quite a bit of figure rearrangement, careful to make color schemes
    more distinguishable.
\end{response}

3) Line 151-152: The closed form approached used here should be contrasted with
the fact that it is in principle not possible to include the effects of other
loci, unlike in these other approaches.

\begin{response}
    Thank you for the suggestion. I now point out that it not possible to include
    the effects of other linked loci above this section, when introducing the
    methods.
\end{response}

@@

4) Figure 4: consider making the y axis consistent between panels.

\begin{response}
    There are new figures in this version of the paper, with hopefully cleaner presentation.
\end{response}

### 2022-03-31 manuscript/paper.tex
ready with the main text

Selected mutations interfere and interact with evolutionary processes at nearby
loci, distorting allele frequency trajectories and creating correlations
between pairs of mutations. A number of recent studies have used patterns of
linkage disequilibrium (LD) between selected variants to test for selective
interference and epistatic interactions, with some disagreement over
interpreting observations from data. Interpretation is hindered by a lack of
analytic or even numerical expectations for patterns of variation between pairs
of loci under the combined effects of selection, dominance, epistasis, and
demography. Here, I develop a numerical approach to compute the expected
two-locus sampling distribution under diploid selection with arbitrary
epistasis and dominance, recombination, and variable population size. I use
this to explore how epistasis and dominance affect expected signed LD,
including for non-steady-state demography relevant to human populations. Using
whole-genome sequencing data from humans, I explore genome-wide patterns of LD
within protein-coding genes. I show that positive LD between missense mutations
within genes is driven by strong positive allele-frequency correlations between
pairs of mutations that fall within the same annotated conserved domain,
pointing to compensatory mutations or antagonistic epistasis as the prevailing
mode of interaction within conserved genic elements. LD between missense
mutations is reduced outside of conserved domains, as would expected under
Hill-Robertson interference. The heterogeneous landscape of both mutational
fitness effects and selective interactions within protein-coding genes calls
for more refined inferences of the joint distribution of fitness and
interactive effects, and the methods presented here should prove useful in that
pursuit.

@@

Most new mutations that affect fitness are deleterious and tend to be
eliminated from a population. The amount of time that a deleterious mutation
segregates depends on the strength of selection against genomes that carry it,
with very damaging mutations kept at low frequencies and purged relatively
rapidly. In the time between mutation and fixation or loss, selected variants,
both beneficial and damaging, can dramatically impact patterns of variation in
nearby linked regions \citep[e.g.][]{Smith1974-em,Charlesworth1995-dq,Kim2000-on}.
This distortion away from neutral expectations has been empirically documented
using sequencing data from an ever-growing set of study systems
\citep{Novembre2009-kc,Cutter2013-mm,Comeron2014-oy}, but questions
remain about the primary mode of interactions between multiple linked variants
and their joint effects on genome-wide patterns of diversity.

In their foundational paper, \citet{Hill1966-gv} recognized that linked
selected variants reciprocally impede the efficacy of selection at each locus,
a process known as selective interference. Linked selection reduces the
fixation probability of advantageous mutations and increases that of
deleterious mutations compared to expectations under single-locus models
\citep{Birky1988-jm}. Allele frequency dynamics and correlations of linked
selected variants are also predicted to deviate from single-locus expectations.
Under a multiplicative fitness model, where the fitness reduction of a genome
carrying multiple deleterious variants is equal to the product of the fitness
reduction of each independent mutation, we expect to see net linkage
disequilibrium (LD) equal to zero for unlinked sites \citep{Kondrashov1995-va}.
But for linked loci, those mutations are expected to segregate on different
haplotypes more often than together, leading to negative, or repulsion, LD,
although the extent of LD depends non-trivially on the strength of selection
and probability of recombination separating loci
\citep{Hill1966-gv,McVean2000-ox}.

@@

In this paper, I develop a numerical approach to solve for the two-locus
sampling distribution under a general diploid selection model with variable
recombination and single-population size history. I use this model to describe
how epistasis and dominance shape expected patterns of signed LD, under both
steady-state and non-equilibrium demography, that have been used to test for
interference and epistasis in population genomic data. I then turn to human
sequencing data and compare patterns of LD for synonymous, missense, and
loss-of-function mutations in protein-coding genes and annotated conserved
domains. I show that while synonymous and missense variants display similar
slightly positive average LD within genes, for missense mutations this signal
is driven by correlations between pairs of mutations within, but not between or
outside of, protein-coding domains. This suggests an importance for
antagonistic (or ``diminishing-returns'') epistasis or a prevalence of
compensatory nonsynonymous mutations within conserved elements.

@@

Directed mutational studies are not possible in most natural populations, and
we must turn to population genetic approaches to infer selective interactions
between observed segregating polymorphisms. Motivated by theory that linked
negatively selected mutations will display negative LD due to interference
\citep{Hill1966-gv}, and that epistasis will drive expected LD away from zero,
a number of recent studies have used patterns of LD within classes of
putatively selected variants to infer modes of selective interactions.
\citet{Callahan2011-ac} observed that pairs of tightly linked nonsynonymous
mutations cluster together more than expected along the same lineages in the
\emph{Drosophilid} species complex, and that those clustered mutations tend to
preserve the charge of the protein and were in positive LD compared to pairs of
synonymous mutations at the same distance. From this, they proposed that
compensatory nonsynonymous variants are regularly tolerated and maintained.
More recently, \citet{Taverner2020-lk} replicated this finding across a diverse
set of genera, showing that such epistatic interactions are important for
protein evolution.

\citet{Sohail2017-zq} observed negative LD between loss-of-function variants in
protein-coding genes (loss-of-function mutations include stop gains and losses,
frameshifts, and other nonsense mutations) in both human and fruit fly
populations. This was interpreted as evidence for widespread synergistic
epistasis between these mutations, in which the fitness reduction of multiple
mutations is greater than the product of that of each individual mutations
independently. Both \citet{Sandler2021-of} and \citet{Garcia2021-zn} have
recently reevaluated patterns of LD between coding variants in humans, fruit
flies, and \emph{Capsella grandiflora}, and suggested interference and
dominance may instead be driving patterns of LD \citep{Garcia2021-zn} or
questioned whether LD between loss-of-function variants is significantly
different from zero \citep{Sandler2021-of}.

A number of factors impede our interpretation of patterns of signed LD between
coding variants. First, for strongly deleterious or loss-of-function mutations,
their low allele frequencies mean that statistical measurement of LD
and other diversity measures are very noisy. Second, comparisons are based on
theory with limiting assumptions, such as steady-state demography, simple
selection and interaction models, or unlinked loci. To generate predictions
under more complex models, we rely on expensive forward simulations. Such
simulations can help build intuition and be used to test inference methods, but
they do not efficiently provide expectations for quantities of interest across
the range of relevant parameters. Analytical and numerical methods for expected
haplotype frequencies and LD under general selective interaction models are
thus crucial for interpreting patterns of variation observed in data.

@@

\begin{figure}[tb!]
    \centering
    \includegraphics{../figures/two-locus-spectrum/stats_summaries}
    \caption{
        \textbf{Sampling distributions and their summaries.}
        Low-order summaries of sampling distributions are commonly computed
        for allele frequencies (A, the site-frequency spectrum, SFS)
        and two-locus haplotype distributions (B, linkage disequilibrium, LD).
        Demographic and selective processes affect both the SFS and LD,
        and observations of nonzero values of Tajima's $D$ or signed LD
        ($\sigma_d^1$) are often taken as evidence for selection or interactions
        between loci, respectively.
        (C) The full two-locus haplotype sampling distribution is a three-dimensional
        object, making it difficult to visualize. We can instead visualize conditional
        distributions of the full sampling distribution, e.g. conditioned on observing
        five copies of the $A$ allele at the left locus and five copies of $B$ at the
        right locus \citep[e.g.][]{Hudson2001-sg}.
        (D) $\sigma_d^2$, which is closely related to $r^2$, is expected to decay
        with increasing recombination distance between loci \citep{Ohta1969-ie}.
        Selection distorts the level of squared LD away from neutral expectations.
        (E) $\sigma_d^1$, refered to as signed LD here, is expected to be zero for
        pairs of neutral mutations \citep{Hill1968-vu}. Interference between linked
        selected mutation causes negative signed LD \citep{Hill1966-gv}, and other
        forms of interactions between selected mutations can cause large negative
        or positive signed LD.
    }
    \label{fig:TwoLocusFS}
\end{figure}

In the Methods, I expand on the moments system developed in
\citet{Ragsdale2019-nt} to compute the expected sampling distribution
\(\Psi_n\) (Figure~\ref{fig:TwoLocusFS}) of two-locus haplotypes under a
general model of selective interactions. This sampling distribution stores the
expected density or observed counts of pairs of biallelic loci with each
possible haplotype frequency configuration in a sample of size \(n\). Below, we
compute expectations for \(\Psi_n\) under varying scenarios of selection and
interaction between pairs of loci. It is not possible in this framework to
include the effects of additional linked selected mutations, such as background
selection due to many linked variants, and individual-based forward simulations
are still needed for such scenarios (e.g.,
Figures~\ref{fig:bgs1}--\ref{fig:bgs4}).

In many cases it is simpler to visualize summaries such as the expectation or
variance of \(D\) (Figure~\ref{fig:TwoLocusFS}D, E) or conditional slices of the
distribution (Figure~\ref{fig:TwoLocusFS}C) instead of the full
three-dimensional distribution. Here I focus on low-order LD statistics
including \(\E[D]\) and \(\E[D^2]\) and their decay with recombination
distance, as these are statistics that are commonly used to test for
interactions between loci. For pairs of biallelic loci, with alleles labeled
\(A/a\) at the left locus and \(B/b\) at the right locus, \(D=f_{AB}-f_A f_B\)
is the standard covariance measure of LD, where \(f_{AB}\) is the frequency of
haplotypes carrying both \(A\) and \(B\), and \(f_A\) and \(f_B\) are the
marginal allele frequencies of \(A\) and \(B\).

@@

\subsection{Clustered mutations may cause positive LD among synonymous mutations}

>>>>

Selected mutations interfere and interact with evolutionary processes at nearby
loci, distorting allele frequency trajectories and creating correlations
between pairs of mutations. A number of recent studies have used patterns of
linkage disequilibrium (LD) between selected variants to test for selective
interference and epistatic interactions, with some disagreement over
interpreting observations from data. Interpretation is hindered by a lack of
analytic or even numerical expectations for patterns of variation between pairs
of loci under the combined effects of selection, dominance, epistasis, and
demography. Here, I develop a numerical approach to compute the expected
two-locus sampling distribution under diploid selection with arbitrary
epistasis and dominance, recombination, and variable population size. I use
this to explore how epistasis and dominance affect expected signed LD,
including for non-steady-state demography relevant to human populations. Using
whole-genome sequencing data from humans, I explore genome-wide patterns of LD
within protein-coding genes. I show that positive LD between missense mutations
within genes is driven by strong positive allele-frequency correlations between
pairs of mutations that fall within the same annotated conserved domain,
pointing to compensatory mutations or antagonistic epistasis as the prevailing
mode of interaction within conserved genic elements. LD between missense
mutations is reduced outside of conserved domains, as would expected under
Hill-Robertson interference. This variation in both mutational fitness effects
and selective interactions within protein-coding genes calls for more refined
inferences of the joint distribution of fitness and interactive effects, and
the methods presented here should prove useful in that pursuit.

@@

Most new mutations that affect fitness are deleterious and tend to be
eliminated from a population. The amount of time that a deleterious mutation
segregates depends on the strength of selection against genomes that carry it,
with very damaging mutations kept at low frequencies and purged relatively
rapidly. In the time between mutation and fixation or loss, selected variants,
both beneficial and damaging, can dramatically impact patterns of variation in
nearby linked regions \citep[e.g.,][]{Smith1974-em,Charlesworth1995-dq,Kim2000-on}.
This distortion away from neutral expectations has been empirically documented
using sequencing data from an ever-growing set of study systems
\citep{Novembre2009-kc,Cutter2013-mm,Comeron2014-oy}, but questions
remain about the primary mode of interactions between multiple linked variants
and their joint effects on genome-wide patterns of diversity.

In their foundational paper, \citet{Hill1966-gv} recognized that linked
selected variants reciprocally impede the efficacy of selection at each locus,
a process known as selective interference. Linked selection reduces the
fixation probability of advantageous mutations and increases that of
deleterious mutations compared to expectations under single-locus models
\citep{Birky1988-jm}. Allele frequency dynamics and correlations of linked
selected variants are also predicted to deviate from single-locus expectations.
Under a multiplicative fitness model, where the fitness reduction of a genome
carrying multiple deleterious variants is equal to the product of the fitness
reduction of each independent mutation, we expect to see net linkage
disequilibrium (LD) equal to zero for unlinked sites \citep{Kondrashov1995-va}.
But for linked loci, those mutations are expected to segregate on different
haplotypes more often than together, leading to negative, or repulsion, LD,
although the extent of LD depends non-trivially on the strength of selection
and the probability of recombination separating loci
\citep{Hill1966-gv,McVean2000-ox}.

@@

In this paper, I develop a numerical approach to solve for the two-locus
sampling distribution under a general diploid selection model with variable
recombination and single-population size history. I use this model to describe
how epistasis and dominance shape expected patterns of signed LD, under both
steady-state and non-equilibrium demography, that have been used to test for
interference and epistasis in population genomic data. I then turn to human
sequencing data and compare patterns of LD for synonymous, missense, and
loss-of-function mutations in protein-coding genes and annotated conserved
domains. I show that while synonymous and missense variants display similar
slightly positive average LD within genes, for missense mutations this signal
is driven by correlations between pairs of mutations within, but not between or
outside of, protein-coding domains. This suggests an importance for
antagonistic epistasis or a prevalence of compensatory nonsynonymous mutations
within conserved elements.

@@

Directed mutational studies are not possible in most natural populations, and
we must turn to population genetic approaches to infer selective interactions
between observed segregating polymorphisms. Motivated by theoretical
predictions that linked negatively selected mutations will display negative LD
due to interference \citep{Hill1966-gv} and that epistasis will drive expected
LD away from zero, a number of recent studies have used patterns of LD within
classes of putatively selected variants to infer modes of selective
interactions.  \citet{Callahan2011-ac} observed that pairs of tightly linked
nonsynonymous mutations cluster together more than expected along the same
lineages in the \emph{Drosophilid} species complex, and that those clustered
mutations tend to preserve the charge of the protein and were in positive LD
compared to pairs of synonymous mutations at the same distance. From this, they
proposed that compensatory nonsynonymous variants are regularly tolerated and
maintained.  More recently, \citet{Taverner2020-lk} replicated this finding
across a diverse set of genera, showing that such epistatic interactions are
important for protein evolution.

\citet{Sohail2017-zq} observed negative LD between loss-of-function variants in
protein-coding genes (loss-of-function mutations include stop gains and losses,
frameshifts, and other nonsense mutations) in both human and fruit fly
populations. This was interpreted as evidence for widespread synergistic
epistasis between these mutations, in which the fitness reduction of multiple
mutations is greater than the product of that of each individual mutation
independently. Both \citet{Sandler2021-of} and \citet{Garcia2021-zn} have
recently reevaluated patterns of LD between coding variants in humans, fruit
flies, and \emph{Capsella grandiflora}, and suggested interference and
dominance may instead be driving patterns of LD \citep{Garcia2021-zn} or
questioned whether LD between loss-of-function variants is significantly
different from zero \citep{Sandler2021-of}.

A number of factors impede our interpretation of patterns of signed LD between
coding variants. First, for strongly deleterious or loss-of-function mutations,
their low allele frequencies mean that measurements of LD and other diversity
statistics are very noisy. Second, comparisons are based on theory with
limiting assumptions, such as steady-state demography, simple selection and
interaction models, or unlinked loci. To generate predictions under more
complex models, we rely on expensive forward simulations. Such simulations can
help build intuition and be used to test inference methods, but they do not
efficiently provide expectations for quantities of interest across the range of
relevant parameters. Analytical and numerical methods for expected haplotype
frequencies and LD under general selective interaction models are thus crucial
for interpreting patterns of variation observed in data.

@@

\begin{figure}[tb!]
    \centering
    \includegraphics{../figures/two-locus-spectrum/stats_summaries}
    \caption{
        \textbf{Sampling distributions and their summaries.}
        Low-order summaries of sampling distributions are commonly computed
        for allele frequencies (A, the site-frequency spectrum, SFS)
        and two-locus haplotype distributions (B, linkage disequilibrium, LD).
        Demographic and selective processes affect both the SFS and LD,
        and observations of nonzero values of Tajima's $D$ or signed LD
        ($\sigma_d^1$) are often taken as evidence for selection or interactions
        between loci, respectively.
        (C) The full two-locus haplotype sampling distribution is a three-dimensional
        object, making it difficult to visualize. We can instead visualize conditional
        distributions of the full sampling distribution, e.g., conditioned on observing
        \(n_A\) copies of the \(A\) allele at the left locus and \(n_B\) copies of
        \(B\) at the right locus \citep[e.g.,][]{Hudson2001-sg}.
        (D) $\sigma_d^2$, which is closely related to $r^2$, is expected to decay
        with increasing recombination distance between loci \citep{Ohta1969-ie}.
        Selection distorts squared LD away from neutral expectations.
        (E) $\sigma_d^1$, refered to as signed LD here, is expected to be zero for
        pairs of neutral mutations \citep{Hill1968-vu}. Interference between linked
        selected mutation causes negative signed LD \citep{Hill1966-gv}, and other
        forms of interactions between selected mutations can cause large negative
        or positive signed LD.
    }
    \label{fig:TwoLocusFS}
\end{figure}

In the Methods, I expand on the moments system developed in
\citet{Ragsdale2019-nt} to compute the expected sampling distribution of
two-locus haplotypes (\(\Psi_n\), Figure~\ref{fig:TwoLocusFS}) under a general
model of selective interactions. This sampling distribution stores the expected
density or observed counts of pairs of biallelic loci with each possible
haplotype frequency configuration in a sample of size \(n\). Below, we compute
expectations for \(\Psi_n\) under varying scenarios of selection and
interaction between pairs of loci. It is not possible in this framework to
include the effects of additional linked selected mutations, such as background
selection due to many linked variants, and individual-based forward simulations
are still needed for such scenarios (e.g.,
Figures~\ref{fig:bgs1}--\ref{fig:bgs4}).

In many cases it is simpler to visualize summaries such as the expectation or
variance of \(D\) (Figure~\ref{fig:TwoLocusFS}D, E) or conditional slices of
the distribution (Figure~\ref{fig:TwoLocusFS}C) instead of the full
three-dimensional distribution \(\Psi_n\). Here I focus on low-order LD
statistics including \(\E[D]\) and \(\E[D^2]\) and their decay with
recombination distance, as these are statistics that are commonly used to test
for interactions between loci. For pairs of biallelic loci, with alleles
labeled \(A/a\) at the left locus and \(B/b\) at the right locus,
\(D=f_{AB}-f_A f_B\) is the standard covariance measure of LD, where \(f_{AB}\)
is the frequency of haplotypes carrying both \(A\) and \(B\), and \(f_A\) and
\(f_B\) are the marginal allele frequencies of \(A\) and \(B\).

@@

\subsection{Causes of positive LD between synonymous mutations}
